---
title: "Predicting Pentecostal Affilitation within Latin America using Subset of the Questionnaire of a Large-Scale Representative Survey"
author: "Gustavo Arruda"
date: "6/4/2021"
output: github_document
---

```{r setup, include=TRUE}

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(haven)
library(tidyverse)
library(psych)
library(MASS)
library(knitr)
set.seed(123456789)

```

# Preparing data

```{r script}

rel_vars <- c('Q28a', 'Q28c', 'Q46a', 'Q55a',
  'Q55b', 'Q55d', 'Q55e', 'Q55f',
  'Q60a', 'Q60e', 'Q60f', 'Q60j')
rel_df <-
  read_sav("Religion in Latin America Dataset.sav",
           col_select = all_of(rel_vars))

rel_tot <- as.data.frame(rel_df) %>%
  zap_label() %>%
  zap_labels() %>%
  mutate_all(funs(as.numeric(str_replace_all(., "2", "0")))) %>%
  na_if(., "98") %>%
  na_if(., "99")

items_tot <- rel_tot[3:12]
rel_tot <- rel_tot %>%
  mutate(score = rowSums(items_tot))

rel_df <- as.data.frame(rel_df) %>%
  zap_missing() %>%
  zap_label() %>%
  zap_labels() %>%
  drop_na() %>%
  mutate_all(funs(as.numeric(str_replace_all(., "2", "0")))) %>%
  na_if(., "98") %>%
  na_if(., "99") %>%
  drop_na()

items <- rel_df[3:12]
rel_df <- rel_df %>%
  mutate(score = rowSums(items))


model_P = lda(Q28a ~ score, data = rel_df)
predictions_P = predict(model_P)
tab_lda_P = table(rel_df$Q28a,predictions_P$class)

cutoff <- ceiling((5.349445 + 6.590104) / 2)

# Cross-Validation, from example on Canvas.
n = nrow(rel_df)
nt = 2000
neval=n-nt
rep=100
hitr=dim(rep)
for(i in 1:rep){
  train=sample(1:n,nt)
  model=lda(Q28a ~ score, data = rel_df[train,])
  predict(model,rel_df[-train,])$class
  tab_lda=table(rel_df$Q28a[-train],
                predict(model,rel_df[-train,])$class)
  hitr[i]=sum(diag(tab_lda))/neval
  }

```

# Objectives

	This project seeks to develop and validate a scale to measure Pentecostal affiliation among Latin American Christians. To accomplish that, I am gathering dichotomous questions from a large-scale survey conducted by the Pew Research Center, chosen by their face validity in relation to the literature on the Pentecostal movement, estimating their composite reliability, calculated with the coefficient alpha, and their criterion validity, estimated with Discriminant Analysis to a self-identification question from the same questionnaire. With such a scale, we could explore Pentecostal affiliation and its political ramifications among different social groups considering more factors than only self-identification.
	
# Literature Review

Contradicting previous sociological views on secularization, which predicted the decline of the importance of religion in the political arena as modernization progressed, since the 1980s religion has re-awakened in the public arena, both in the United States and in Brazil (Casanova 1994). In the United States, the Evangelical and Fundamentalist revival characterized the New Religious Right, backing the Moral Majority, prompting continued analytical relevance within sociology and political science (Hackett and Lindsay 2008; Kellstedt and Smidt 1991). Yet, debates over the appropriate way to identify and measure alignment and identification with Evangelicalism or Protestant Fundamentalism persist, which partially stems from a broader difficulty in establishing appropriate measures of religious identification (Hackett 2014; Hackett and Lindsay 2008; Kellstedt and Smidt 1991; Smith 1990; Steensland, Park et al 2000). The strategy often revolves around using surveys to calculate correlations between Evangelical or Fundamentalist self-identification with scales of attitudes and behaviors theoretically associated with these religious movements (Hackett and Lindsay 2008; Kellstedt and Smidt 1991; Smith 1990; Steensland, Park et al 2000).

Meanwhile, in Brazil, the Pentecostal movement, which shares multiple characteristics with the American Evangelical and Fundamentalists, has presented extraordinary growth since mid-20th century. In many countries, the Pentecostal movement ended up characterized by support to the military dictatorships of the 1970s and 1980s (Chesnut 1997; Martin 1990; Stoll 1990). More recently, in Brazil they now represent a third of the population, with proportional representation in Congress and increasing weight in presidential elections (Nicolau 2014). In the 2018 election of the authoritarian populist President Jair Bolsonaro, who won with 55%, Pentecostals disproportionally voted for him when compared to Catholics and the religiously unaffiliated. By using the “Religion in Latin America” 2014 survey conducted by the Pew Research Center, we have the opportunity to extend the exploration of methodological nuances in measuring denominational affiliation among Christian groups, both by self-identification and attitudinal scales, from the United States to other social contexts like Latin America. That would allow us to refine our knowledge on the political influence of the Pentecostal movement, and its reverberations to global democracy.

# Methods

The data used in this project comes from the “Religion in Latin America” 2014 survey published by the Pew Research Center, a nonpartisan think tank that researches public opinion with an expertise in religion. The survey, through 3 different languages (Spanish, Portuguese and Guarani), included nationally representative samples from 18 different countries, which together include more than 95% of the population in Latin America. In total, the dataset included 30,326 observations, going to 20,221 observations after removing all empty responses from the variables of interest.

For my dependent variable I selected a dichotomous item indicating self-identification with Pentecostalism, along with 10 other dichotomous variables including attitudes and behaviors pertaining to Pentecostal theology. I then created a composite score with the 10 items, which presented face validity according to the literature on the theme. The question asking whether the participant self-identifies as Pentecostal was only applied to Latin American participants already self-identifying as Christians. That means that my scale, if valid, would only have its validity measured to differentiate participant’s religious attitudes and behaviors among strands of Christianism within Latin America. Religious identification might change drastically among cultures, even within a single country (Hackett 2014; Steensland, Park et al 2000), which would require this scale to be evaluated for other populations for their results to be validated.

After that, I calculated the coefficient alpha as a measure of reliability and item statistics of the 10-item composite test, using the function “alpha” in the R package “psych”. The coefficient alpha estimates internal consistency, showing whether the different items in the scale measure statistically correlated underlying constructs, whichever they may be. That, on its own hand, helps us to estimate the reliability of scale items in measuring our construct of interest.

At last, to evaluate the fit of a scale composed by the 10 items of the survey to classify and predict religious affiliation with Pentecostalism I used Discriminant Analysis (DA), using the function “lda” from the R package “MASS”. Two sub-groups — Non-Pentecostals and Pentecostals — were established according to a self-identification question included in the same survey. The DA model created a linear function that, when applied to each individual test results, creates a discriminant score that maximizes the difference between its sub-group means. The average between the two sub-group means characterizes the cut-off value of the DA model, which is used to classify each score to the Non-Pentecostal or Pentecostal group depending on its standing above or below the cut-off. I evaluated the fit of the model by cross-validating it, which entails applying it to various random subsamples of the data, before verifying the proportion of cases that it predicts correctly, also named hit ratio. The linear function calculated by the DA model ultimately helps us to understand the strength of the relationship between the composite score of the scale I proposed and the self-identification measure. Discriminant Analysis is most often used to verify criterion validity with multiple predictors, which is not the case here given that I am using a single composite score as the independent variable. Still, I believe that DA might still be a good estimative of criterion validity for my purposes given I am trying to differentiate between two categorial dependent variables.

# Results

When looking at the output of the reliability analysis I conducted in R, we see that the coefficient alpha of the 10-item scale I constructed, pulling from the questionnaire written by the Pew Research for the 2014 “Religion in Latin America” survey, was 0.68. The alpha coefficient is the average of all the possible split-half reliabilities, which is constructed from the covariance of all possible item pairs. The reliability of a scale can also be interpreted as its correlation with itself. Additionally, the item statistics show that dropping each of all the items would not increase the remaining reliability, which is a good indicator that none of the individual items is substantially decreasing the total coefficient alpha. Conversely, dropping each of all the items would also not decrease the remaining reliability, which shows that all of them have relatively similar item total reliabilities, meaning the correlation between each item with the entire scale.

```{r reliability}

print(alpha(items_tot, cumulative = TRUE))

```

The hit ratio was moderately predictive, giving 77.23% of correct predictions, relatively higher than the 50% expected hit ratio. Among 100 Christian participants, about 77% would be appropriately identified as associated with the Pentecostal movement or not. When looking at the confusion table, we see that the ratio of false negatives is much higher than the ratio of false positives, around 40% for the first versus 21% for the second. That means that the scale is much more efficient in distinguishing who is not affiliated with Pentecostalism than who is affiliated with it. The cut-off value is 6, which would be the score to classify affiliation with the Pentecostal movement among Christian individuals in Latin America.

```{r discriminant}

print(tab_lda_P)
print(sum(diag(tab_lda_P))/nrow(rel_df))
print(model_P)
paste('Cutoff: ', cutoff)
print(mean(hitr))

```

# Discussion

Overall, the scale of Pentecostal affiliation that I proposed presents moderate reliability and moderate criterion validity. However, it is important to consider the broad range of populations included in the validation of this scale when evaluating its usefulness— 18 nationally representative surveys spanning 3 different languages—, which means that it moderately measures the construct of Pentecostal affiliation among various populations and contexts. Although this scale presents moderate predictive power, and seems to include moderately predictive individual items, I would suggest as next steps the development and testing of additional items as a way to improve the overall scale performance in reliability and validity measures, particularly to diminish its false-negative rates. Another venue of further exploration would be testing the reliability and validity of this scale for more specific populations, given that it possibly holds more significant classificatory power for some social groups than others. On the downside, it is important to notice that the usage of self-identification as a criterion for construct validity remains contested, given the complexity of identification and affiliation in regards to a religious movement. For instance, changes in wording result in significant differences in interpretation for different populations (Hackett 2014; Hackett and Lindsay 2008).

# References

Casanova, Jose. 1994. Public Religions in the Modern World. 2nd ed. Chicago, IL: University of Chicago Press.

Chesnut, R. Andrew. 1997. Born Again in Brazil: The Pentecostal Boom and the Pathogens of Poverty. 1st Editio. New Brunswick: Rutgers University Press.

Hackett, Conrad. 2014. “Seven Things to Consider When Measuring Religious Identity.” Religion 44(3):396–413. doi: 10.1080/0048721X.2014.903647.

HACKETT, CONRAD, and D. MICHAEL LINDSAY. 2008. “Measuring Evangelicalism: Consequences of Different Operationalization Strategies.” Journal for the Scientific Study of Religion 47(3):499–514. doi: https://doi.org/10.1111/j.1468-5906.2008.00423.x.

Kellstedt, Lyman, and Corwin Smidt. 1991. “Measuring Fundamentalism: An Analysis of Different Operational Strategies.” Journal for the Scientific Study of Religion 30(3):259–78. doi: 10.2307/1386972.

Martin, David. 1990. Tongues of Fire: Explosion of Protestantism in Latin America. London, England: Blackwell.

Nicolau, Jairo. 2014. “Determinantes Do Voto No Primeiro Turno Das Eleições Presidenciais Brasileiras de 2010: Uma Análise Exploratória.” Opiniao Publica 20(3):311–25. doi: 10.1590/1807-01912014203311.

Pew Research Center. 2014. “Appendix A: Methodology” Religion in Latin America. Retrieved (https://www.pewforum.org/2014/11/13/appendix-a-methodology/).
Smith, Tom W. 1990. “Classifying Protestant Denominations.” Review of Religious Research 31(3):225–45. doi: 10.2307/3511614.

Steensland, Brian, Jerry Z. Park, Mark D. Regnerus, Lynn D. Robinson, W. Bradford Wilcox, and Robert D. Woodberry. 2000.

“The Measure of American Religion: Toward Improving the State of the Art.” Social Forces 79(1):291–318. doi: 10.2307/2675572.

Stoll, David. 1990. Is Latin America Turning Protestant?: The Politics of Evangelical Growth. Berkeley, CA: University of California Press.

# Appendix A – Test Items

## criterion VARIABLE (PENTECOSTAL SELF-IDENTIFICATION)

 - ASK IF CHRISTIAN (QCURREL = 1,2 OR QCURRELc=1) (IF CHRISTIAN, INCLUDING CATHOLICS AND ALL SELF-IDENTIFIED CHRISTIANS)
Q28	Would you describe yourself as a (INSERT ITEM), or not? And would you describe
yourself as a (INSERT ITEM) or not? 

a.	pentecostal 
b. 	NO ITEM B
c.	charismatic 
ASK IF PROTESTANT (CURREL = 2)
d.	“born again” 

1	Yes
2	No
98	Don’t know (DO NOT READ)
99	Refused (DO NOT READ)

## CONSTRUCT/PREDICTOR VARIABLES (PENTECOSTAL ATTITUDES AND BEHAVIORS)

### DICHOTOMOUS (10 QUESTIONS)

 - ASK ALL
Q55	Have you ever (INSERT) 

a.	experienced or witnessed a divine healing of an illness or injury? 
b.	given or interpreted prophecy? 
c.	NO ITEM C
d.	received a direct revelation from God? 
e.	experienced or witnessed the devil or evil spirits being driven out of a person? 
f.	spoken or prayed in tongues? 

1	Yes
2	No
98	Don’t know (DO NOT READ)
99	Refused (DO NOT READ)

 - ASK ALL
Q60	Which, if any, of the following do you believe? Do you believe (INSERT), or not?
 (READ LIST) 

a.	in God? 
e.	in Heaven? 
f.	in Hell? 
j.	in Miracles? 

1	Yes
2	No
98	Don’t know (DO NOT READ)
99	Refused (DO NOT READ)

 - ASK IF ATTENDS RELIGIOUS SERVICES MORE THAN NEVER (Q45 = 1 THRU 5)
Q46	And does the church or house of worship where you most often attend religious services (INSERT)? (READ LIST) 
	
a.	Try to bring others into the faith by spreading the message of Christ? 
	
1	Yes
2	No
98	Don’t know (DO NOT READ)
99	Refused (DO NOT READ)


